---
title: "Program Code_CAnD3 RRWM Exercise"
author: "Leah Houseman"
date: "`r Sys.Date()`"
output:
   pdf_document: default
   html_document:
     keep_md: yes
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

#packages

library(tidyverse)
library(tidymodels)
library(knitr)
library(kableExtra)
library(gt)
library(broom)
library(ggplot2)
library(naniar)
```

#### **Download data and call in dataset**
- Dataset is available to download from Canvas course for CAnD3 fellows only. For more information on how to access this data for purposes beyond the CAnD3 Training Program, see the `Data Accessibility Statement` in this repository. 
- Call in dataset **(Be sure to change the file location to match the dataset on your local computer & call the dataframe "data" in order to ensure the code works without editing)**
```{r load data}
data <- read_csv("/Users/leahj/Desktop/CAnD3/CAnD3/CAnD3_RWork/RRWM Exercise/CAnD3-RRWM-Exercise/Dataset_gss-12M0025-E-2017-c-31_F1.csv")
```

##### **Take initial look at data to examine structure**
```{r view data}
glimpse(data)
```

#### **Understand the nature of missing data**
```{r evaluate missing}
missing_table <- miss_var_summary(data)
missing_table
```
- We can see that the missing data has been accounted for via special missing values. These will need to be dealt with for the variables of interest. 

#### **Define Research Question and Recode Key Variables**
- This analysis will aim to answer the questions 1) What percentage of Canadians aged between 14-49 intend to have a child in the next 3 years? and 2) Does this vary by self-rated physical health? 
- The sample will be limited to those aged under 50 (AGEC) and who (or whose spouse) have not been told that they cannot have any(more) biological children (Respondent = FI_140, Spouse = FI_240: 1 = Yes, 2 = No).
- The primary independent variable (CH3YRS) will use a survey question (FI_105) that asked "do you intend to have a(nother) child within the next 3 years?
- Drop missing data from variables of interest. Usually, a data imputation method would be used to handle missing data, but for the purposes of this exercise, drop the missing values from the dataframe.

**Remove those over age 50 and those who or (whose spouse) have been told they cannot have any(more) biological children**

```{r remove over 50 & infertile}
#Remove respondents over age 50
glimpse(data$AGEC) #view variable
summary(data$AGEC)
#Remove those over age 50
data <- data |> subset(AGEC < 50)
#View variable after change
glimpse(data$AGEC)
summary(data$AGEC)

#Exclude if respondent or spouse has been told they cannot have any(more) bio children. 1 = Has bee told, 2 = Has not been told. 6-9 represent different flavours of missing data, so they will be assigned as missing and dropped from this dataframe. 
glimpse(data$FI_140)
summary(data$FI_140)
#Assign missing values (6-9)
data <- data |> 
  replace_with_na(replace = list(FI_140 = c(6, 7, 8, 9)))
#Remove those who responded yes
data <- data |> subset(FI_140 != 1)
#View variable to ensure this worked as intended
summary(data$FI_140)
table(data$FI_140)


#Do the same for the spouse (FI_240)
glimpse(data$FI_240)
summary(data$FI_240)
#Assign missing values (6-9)
data <- data |> 
  replace_with_na(replace = list(FI_240 = c(6, 7, 8, 9)))
#Remove those who responded yes
data <- data |> subset(FI_240 != 1)
#View variable to ensure this worked as intended
summary(data$FI_240)
table(data$FI_240)
```

**Create a binary outcome variable, "CH3YRS" to reflect fertility intentions (Yes/No) within the next 3 years**
```{r create outcome var}
#View variable to examine structure
glimpse(data$FI_105)
summary(data$FI_105)
table(data$FI_105)

#We can tell based on the codebook that 1 = Definitely yes, 2 = Probably yes, 3 = Probably Not, 4 = No, definitely not, and 5 = unsure. 6-9 represent different flavours of missing data, so they will be assigned as missing and dropped from this dataframe. 

data <- data |> 
  replace_with_na(replace = list(FI_105 = c(6, 7, 8, 9)))
summary(data$FI_105) #view variable to ensure this worked

#Now I will transform this variable into yes/no variable, CH3YRS
data <- data |> 
  mutate(CH3YRS = case_when(
    FI_105 == 1 ~ "1_Yes",
    FI_105 == 2 ~ "1_Yes", 
    FI_105 == 3 ~ "2_No",
    FI_105 == 4 ~ "2_No",
    FI_105 == 5 ~ "2_No",
    FI_105 == 6 ~ "2_No",
  ))
table(data$FI_105, data$CH3YRS) #compare old and new variable to ensure recode worked as intended
#view new variable to ensure it looks as intended
glimpse(data$CH3YRS)
data$CH3YRS <- as.factor(data$CH3YRS) #recode as factor variable
summary(data$CH3YRS)
table(data$CH3YRS)

data$FI_105 <- NULL #Remove old variable
```

**Recode the primary independent variable, Self-Rated Physical Health (SRPH using codebook variable SRH_110)**

```{r recode input var}
#View variable
glimpse(data$SRH_110)
summary(data$SRH_110)
table(data$SRH_110)
#The "Don't know," "refusal," and "not stated" responses (6-9) need to be designated as missing
data <- data |> 
  replace_with_na(replace = list(SRH_110 = c(6, 7, 8, 9)))
summary(data$SRH_110) #view variable to ensure this worked
#This variable is originally coded as 5 = poor and 1 = excellent, but for ease of interpretation, I will recode the scale to start with poor = 1 and end with excellent = 5.
data <- data |> 
  mutate(SRPH = case_when(
    SRH_110 == 1 ~ "5_Excellent",
    SRH_110 == 2 ~ "4_Very Good",
    SRH_110 == 3 ~ "3_Good",
    SRH_110 == 4 ~ "2_Fair",
    SRH_110 == 5 ~ "1_Poor",
  ))

glimpse(data$SRPH) # View recoded variable
data$SRPH <- as.factor(data$SRPH) #Recode as  factor variable

#View old variable and new to make sure recode processed correctly, view new variable
table(data$SRH_110, data$SRPH)
glimpse(data$SRPH)
summary(data$SRPH)

data$SRH_110 <- NULL #Remove old variable
```

**Age, measured in whole numbers (Codebook variable AGEC)**
```{r recode age}
#Since there are multiple measures of age (Groups of 5/10-year intervals, a whole number version, and a decimal point version), I am going to keep the one that measures in whole numbers (AGEC). The rest can be made null.
data$AGEDC <- NULL
data$AGEGR5 <- NULL
data$AGEGR10 <- NULL

#View the variable to confirm it does not need recoding
glimpse(data$AGEC)
summary(data$AGEC)
table(data$AGEC) #There are no missing data on this one.
```

***Sex***
```{r recode sex}
#View variable
glimpse(data$SEX)
summary(data$SEX)
table(data$SEX) #There are no missing data on this variable. 

#For ease of interpretation, I am adding labels to this variable.
data <- data |> 
  mutate(SEX = case_when(
    SEX == 1 ~ "1_Male",
    SEX == 2 ~ "2_Female"
  ))
data$SEX <- as.factor(data$SEX) #recode as factor variable
glimpse(data$SEX) # View recoded variable
summary(data$SEX)
```
  
***Martial Status***

```{r recode marstat}
#View variable
glimpse(data$MARSTAT) 
summary(data$MARSTAT)
table(data$MARSTAT) #There are no remaining missing data after we removed those out of scope of this analysis.

#This variable has multiple categories that effectually mean "no longer married" for various reasons. For ease of analysis, I am going to combine these categories.
data <- data |> 
  mutate(MARSTAT = case_when(
    MARSTAT == 1 ~ "1_Married",
    MARSTAT == 2 ~ "2_Common Law",
    MARSTAT == 3 ~ "3_Widowed/Separated/Divorced",
    MARSTAT == 4 ~ "3_Widowed/Separated/Divorced",
    MARSTAT == 5 ~ "3_Widowed/Separated/Divorced",
    MARSTAT == 6 ~ "4_Single",
  ))
data$MARSTAT <- as.factor(data$MARSTAT) #recode as factor variable
glimpse(data$MARSTAT) # View recoded variable
summary(data$MARSTAT)
```

***Education level (EDULVL, using codebook variable EHG3_01B)***
```{r recode edulvl}
#View variable
glimpse(data$EHG3_01B) 
summary(data$EHG3_01B)
table(data$EHG3_01B)
#The "Don't know," "refusal," and "not stated" responses (96-99) need to be accounted as missing
data <- data %>%
  replace_with_na(replace = list(EHG3_01B = c(96, 97, 98, 99)))
summary(data$EHG3_01B) #check to make sure this worked

#This variable has multiple categories that effectually mean "post-secondary education below the bachelor's level". For ease of analysis, I am going to combine some of these categories and add labels
data <- data |> 
  mutate(EDULVL = case_when(
    EHG3_01B == 1 ~ "1_Less than H.S.",
    EHG3_01B == 2 ~ "2_H.S/Equiv",
    EHG3_01B == 3 ~ "3_P.S under Bach",
    EHG3_01B == 4 ~ "3_P.S under Bach",
    EHG3_01B == 5 ~ "3_P.S under Bach",
    EHG3_01B == 6 ~ "4_Bach Degree",
    EHG3_01B == 7 ~ "5_Post Bach Lvl",
  ))

glimpse(data$EDULVL) # View recoded variable
data$EDULVL <- as.factor(data$EDULVL) #Recode as factor variable

#View old variable and new to make sure recode processed correctly, view new variable
table(data$EHG3_01B, data$EDULVL)
glimpse(data$EDULVL)
summary(data$EDULVL)

data$EHG3_01B <- NULL #Remove old variable 
```


**Drop missing values from variables of interest (CH3YRS, SRPH, AGEC, SEX, MARSTAT, EDULVL)**
```{r}
data <- data |> 
  filter(
    !is.na(CH3YRS),
    !is.na(SRPH),
    !is.na(AGEC),
    !is.na(SEX),
    !is.na(MARSTAT),
    !is.na(EDULVL),
  )

#View variables to ensure this worked
summary(data$CH3YRS)
summary(data$SRPH)
summary(data$AGEC)
summary(data$MARSTAT)
summary(data$EDULVL)

```

**Create a table to display the percent and count for the CH3YRS variable by EDULVL variable, called "edulvl_table**
```{r}
edulvl_table <- data |> 
  count(CH3YRS, EDULVL) |> 
  group_by(CH3YRS) |> 
  mutate(percent = round(100 * n / sum(n), 1)) |> 
  ungroup() |> 
  select(-n) |> 
  pivot_wider(names_from = CH3YRS, values_from = percent) |> 
  gt() |> 
  tab_header(
    title = "Descriptive Statistics by Fertility Intention by Education Level", 
    subtitle = "Percent distribution"
  ) |> 
  cols_label(
    EDULVL = "Education Level", 
    "1_Yes" = "Yes", 
    "2_No" = "No"
  ) |> 
  tab_options(
    table.background.color = "#4D4E53",
  ) |> 
  tab_footnote(md("*Data from 2017 GSS, Statistics Canada
                  N = 4,164*"))

edulvl_table

#Save table
gtsave(edulvl_table, "/Users/leahj/Desktop/CAnD3/CAnD3/CAnD3_RWork/RRWM Exercise/CAnD3-RRWM-Exercise/edulvl_table.html")

```

**Logistic Regression (Output = CH3YRS, Input = SRPH, SEX, EDULVL, MARSTAT, AGEC) with table displaying results, seed set at 567 for replicability purposes, capture session info**
```{r}
set.seed(567) #set the seed for replicability purposes


# Run logistic regression
LogRegressionModel <- glm(CH3YRS ~ SRPH + SEX + EDULVL + MARSTAT + AGEC,
             data = data,
             family = binomial)

# Tidy the model output
LogR_table <- tidy(LogRegressionModel, exponentiate = TRUE, conf.int = TRUE)

# Display results in a formatted table
LogR_table <- LogR_table |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  rename(
    Term = term,
    OR = estimate,
    `CI Lower` = conf.low,
    `CI Upper` = conf.high,
    `P-value` = p.value
  ) |>
  gt() |>
  tab_header(
    title = "Logistic Regression Results",
    subtitle = "Predicting Fertility Intentions"
  ) |> 
  tab_options(
    table.background.color = "#4D4E53",
  ) |> 
  tab_footnote(md("*Data from 2017 GSS, Statistics Canada
                  N = 4,164*"))
LogR_table

#save table
gtsave(LogR_table, "/Users/leahj/Desktop/CAnD3/CAnD3/CAnD3_RWork/RRWM Exercise/CAnD3-RRWM-Exercise/LogR_table.html")

```


```{r}
#Document session info
sessionInfo()
#Save to a file
writeLines(capture.output(sessionInfo()), "session_info.txt")

```
